---
title: "MDS comparisons"
output: pdf_document
---

Comparison on various data sets

data: 
- ABR curves, frequency 100, SPL 70, Genes (everything fixed strong curves)
- ABR curves, frequency 30000, SPL 55, Genes (everything fixed weak curves)
- ABR curves, frequency 12000, SPL 50 - 80 (freq fixed)
- ABR curves, SPL 70 (SPL fixed)
- ABR curves, frequency 12 000 - 30000 SPL 70 (SPL fixed, no freq 100/6000)
- ABR curves, frequency 12000 - 30000, SPL 50 - 80 (similar curves)
- ABR curves (nothing fixed)
- ABR curves with 5% outliers


prep
```{r}
library(ggplot2)
library(dplyr)
library(umap)
library(tidyverse)
library(data.table)
library(tidyfun)
library(checkmate)
library(Rtsne)

source(here::here("Apps/Embed_it/R/utils.R"))

set.seed(42)

```


data preparation

```{r}
prep_curves_comp <- function(curves, mouse_data, freq = c(100, 6000, 12000, 18000, 24000, 30000), SPL = c(10:90), saveit = FALSE, filename = "test") {
  
  mouse_ids <- mouse_data %>% .["mouse_id"]
  curves_temp <- curves[which(curves$mouse_id %in% as.integer(mouse_ids$mouse_id)), ]
  curves_info <- curves_temp %>% filter(frequency %in% freq & sound_level %in% SPL)

  # the same mouse appears multiple times in mouse_data, because it is found in various cohorts
  labels <- mouse_data %>% filter(mouse_id %in% curves_info$mouse_id) #%>% .$cohort_type
  labels <- labels[!duplicated(labels$mouse_id), ] # delete duplicated ids
  labels <- cbind("mouse_id" = labels$mouse_id, "label" = labels$gene_symbol)
  
  functs <- curves_info[ , 1:666]
  functs <- functs[ , -c(4:24)]
  functs_tf <- functs
  
  functs_tf <- functs_tf %>% janitor::clean_names() %>% tf_gather(key = "tf_fun", starts_with("t"))
  
  # combine to dataframe without embeddings
  functs_df_new <- merge(functs_tf, as.data.frame(labels), by = "mouse_id")
  functs_df_new[is.na(functs_df_new)] <- "wild_type"

  functs_df_new
    
}
```


```{r}
# ABR curves, frequency 100, SPL 70, Genes (everything fixed strong curves)
mice_100_70 <- prep_curves_comp(curves20000, mouse_data, freq = 100, SPL = 70)
# 216 obs (would be 130 obs. no wildtype)

# ABR curves, frequency 30000, SPL 45 (everything fixed weak curves)
mice_30000_55 <- prep_curves_comp(curves20000, mouse_data, freq = 30000, SPL = 55)
# 118 obs

# ABR curves, frequency 12000, SPL 50 - 80 (freq fixed

mice_12000_50to80 <- prep_curves_comp(curves20000, mouse_data, freq = 12000, SPL = c(50:80))
# 1082 obs

# ABR curves, SPL 70 (SPL fixed)
mice_70 <- prep_curves_comp(curves20000, mouse_data, SPL = 70)
# 1231 obs

# ABR curves, frequency 12 000 - 30000 SPL 70 (SPL fixed, no freq 100/6000)
mice_12000to30000_70 <- prep_curves_comp(curves20000, mouse_data, freq = c(12000, 18000, 24000, 30000), SPL = 70)
# 798 obs

# ABR curves, frequency 12000 - 30000, SPL 50 - 80 (similar curves)
mice_12000to30000_50to80 <- prep_curves_comp(curves20000, mouse_data, freq = c(12000, 18000, 24000, 30000), SPL = c(50:80))
# 4159 obs

# ABR curves (nothing fixed)
mice_all <- prep_curves_comp(curves20000, mouse_data)
# 11311 obs

# ABR curves with 5% outliers (different freq)
mice_outl <- prep_curves_comp(curves20000, mouse_data, freq = 100, SPL = 70)
mice_outl_5 <- prep_curves_comp(curves20000, mouse_data, freq = 12000, SPL = 70)[1:(nrow(prep_curves_comp(curves20000, mouse_data, freq = 12000, SPL = 70)) * 0.05), ]

mice_outl <- rbind(mice_outl_5, mice_outl)

# combine
datasets_mice <- list("mice_100_70" = mice_100_70, "mice_30000_55" = mice_30000_55,"mice_12000_50to80" = mice_12000_50to80, "mice_70" = mice_70, "mice_12000to30000_70" = mice_12000to30000_70, "mice_12000to30000_50to80" = mice_12000to30000_50to80, "mice_all" = mice_all, "mice_outl" = mice_outl)

```


get the embeddings with different hyperparameter settings 2D

```{r}
# default settings

set.seed(42)
tsne_default <- map(datasets_mice_sm, function(x){ #datasets_mice_sm
  embedding_tsne <- Rtsne(as.matrix(tf_spread(as.data.frame(x$tf_fun))), dims = 2)
  colnames(embedding_tsne$Y) <- paste0("tsne_default_", 1:ncol(embedding_tsne$Y))
  
  # add the labels
  embedding_tsne <- c(embedding_tsne, var = list(x$label))
  
  embedding_tsne
})

```

build datasets for the app

```{r}

# combine the embeddings with the dataset
datasets_mice_app <- datasets_mice
tsne_temp <- tsne_default

datasets_mice_app <- map2(datasets_mice_app, tsne_temp, ~ cbind(.x, .y$Y))

# build app input
datasets_mice_app_input <- map(datasets_mice_app, buildInput, embs_cols = c("tsne_default_1", "tsne_default_2"), vars_cols = c(1,2,3,5), vars_cat_cols = c(1,2,3,5), tf_col = 4, id_col = NULL,
                       img = FALSE, url_col = NULL)

```

Add more embeddings

skip PCA step
```{r}
set.seed(42)
tsne_noPCA <- map(datasets_mice_sm, function(x){
  embedding_tsne <- Rtsne(as.matrix(tf_spread(as.data.frame(x$tf_fun))), dims = 2, pca = FALSE)
  
  colnames(embedding_tsne$Y) <- paste0("tsne_noPCA_", 1:ncol(embedding_tsne$Y))
  
  embedding_tsne
})

datasets_mice_app <- map2(datasets_mice_app, tsne_noPCA, ~ cbind(.x, .y$Y))

```

vary perplexity

```{r}
set.seed(42)
tsne_perp10 <- map(datasets_mice_sm, function(x){
  embedding_tsne <- Rtsne(as.matrix(tf_spread(as.data.frame(x$tf_fun))), dims = 2, perplexity = 10)
  
  colnames(embedding_tsne$Y) <- paste0("tsne_perp10_", 1:ncol(embedding_tsne$Y))
  
  embedding_tsne
})

datasets_mice_app <- map2(datasets_mice_app, tsne_perp10, ~ cbind(.x, .y$Y))

```

# this seems to work really well for freq, especially in mice_70
# although, now that I'm comparing it, not that much of a difference


no Barnes-Hut

```{r}

set.seed(42)
tsne_noBH <- map(datasets_mice_sm, function(x){
  embedding_tsne <- Rtsne(as.matrix(tf_spread(as.data.frame(x$tf_fun))), dims = 2, theta = 0.0)
  
  colnames(embedding_tsne$Y) <- paste0("tsne_noBH_", 1:ncol(embedding_tsne$Y))
  
  
  embedding_tsne
})

datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
```


final step
```{r}
# build app input
datasets_mice_app_input <- map(datasets_mice_app, buildInput, embs_cols = c("tsne_default_1", "tsne_default_2", "tsne_noBH_1", "tsne_noBH_2", "tsne_perp10_1", "tsne_perp10_2", "tsne_noPCA_1", "tsne_noPCA_2"), vars_cols = c(1,2,3,5), vars_cat_cols = c(1,2,3,5), tf_col = 4, id_col = NULL,
                       img = FALSE, url_col = NULL)

```

---------------------------------
reduce the size of the curves

```{r}
# get every fifth value
fifth <-function(x) subset(x, x %% 5 == 0)


get_fifths <- function(x) {
 # browser()
  curve <- tf_spread(x[ , "tf_fun", drop = FALSE])
  #curve <- curve[ , 14:ncol(curve)]
  columns5 <- fifth(1:ncol(curve))
  curve <- curve[ , columns5]
  curve <- tf_gather(curve)
  x$tf_fun <- curve$tf_fun
  x
}
```

```{r}
datasets_mice_app_input1 <- map(datasets_mice_app_input, get_fifths)

```

t-SNE 3D

```{r}
start_time <- Sys.time()
set.seed(42)
tsne_default_3D <- map(datasets_mice_app_input, function(x){
  embedding_tsne <- Rtsne(as.matrix(tf_spread(as.data.frame(x$tf_fun))), dims = 3)
  
  colnames(embedding_tsne$Y) <- paste0("embs_tsne_default3D_", 1:ncol(embedding_tsne$Y))
  
  
  embedding_tsne
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, tsne_default_3D, ~ cbind(.x, .y$Y))

end_time <- Sys.time()
end_time - start_time


```
3.44 minutes
3.52
3.53


Umap

```{r}
start_time <- Sys.time()
set.seed(42)
umap_default <- map(datasets_mice_app_input, function(x){
  
  custom.umap <- umap.defaults
  custom.umap$verbose <- TRUE
  custom.umap$input <- "data"
  
  embedding_umap <- umap(as.matrix(tf_spread(as.data.frame(x$tf_fun))), config = custom.umap)#
  colnames(embedding_umap$layout) <- paste0("embs_umap_default_", 1:ncol(embedding_umap$layout))
  
  embedding_umap
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, umap_default, ~ cbind(.x, .y$layout))

end_time <- Sys.time()
end_time - start_time

```

2.35 minutes
2.31

umap n_neighbors = 3
```{r}
start_time <- Sys.time()
set.seed(42)
umap_k3 <- map(datasets_mice_app_input, function(x){
  
  custom.umap <- umap.defaults
  custom.umap$verbose <- TRUE
  custom.umap$input <- "data"
  custom.umap$n_neighbors <- 3
  
  embedding_umap <- umap(as.matrix(tf_spread(as.data.frame(x$tf_fun))), config = custom.umap)#
  colnames(embedding_umap$layout) <- paste0("embs_umap_k3_", 1:ncol(embedding_umap$layout))
  
  embedding_umap
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, umap_k3, ~ cbind(.x, .y$layout))

end_time <- Sys.time()
end_time - start_time
```

48 seconds


umap n_neighbors = 50
```{r}
start_time <- Sys.time()
set.seed(42)
umap_k50 <- map(datasets_mice_app_input, function(x){
  
  custom.umap <- umap.defaults
  custom.umap$verbose <- TRUE
  custom.umap$input <- "data"
  custom.umap$n_neighbors <- 50
  
  embedding_umap <- umap(as.matrix(tf_spread(as.data.frame(x$tf_fun))), config = custom.umap)#
  colnames(embedding_umap$layout) <- paste0("embs_umap_k50_", 1:ncol(embedding_umap$layout))
  
  embedding_umap
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, umap_k50, ~ cbind(.x, .y$layout))

end_time <- Sys.time()
end_time - start_time
```

6.57 minutes


umap random initialization
```{r}
start_time <- Sys.time()
set.seed(42)
umap_rand <- map(datasets_mice_app_input, function(x){
  
  custom.umap <- umap.defaults
  custom.umap$verbose <- TRUE
  custom.umap$input <- "data"
  custom.umap$init <- "random"
  
  embedding_umap <- umap(as.matrix(tf_spread(as.data.frame(x$tf_fun))), config = custom.umap)#
  colnames(embedding_umap$layout) <- paste0("embs_umap-randinit_", 1:ncol(embedding_umap$layout))
  
  embedding_umap
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, umap_rand, ~ cbind(.x, .y$layout))

end_time <- Sys.time()
end_time - start_time
```
~ 3 minutes

umap min_dist = 0.9
```{r}
start_time <- Sys.time()
set.seed(42)
umap_mindist90 <- map(datasets_mice_app_input, function(x){
  
  custom.umap <- umap.defaults
  custom.umap$verbose <- TRUE
  custom.umap$input <- "data"
  custom.umap$min_dist <- 0.9
  
  embedding_umap <- umap(as.matrix(tf_spread(as.data.frame(x$tf_fun))), config = custom.umap)#
  colnames(embedding_umap$layout) <- paste0("embs_umap-mindist0.9_", 1:ncol(embedding_umap$layout))
  
  embedding_umap
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, umap_mindist90, ~ cbind(.x, .y$layout))

end_time <- Sys.time()
end_time - start_time
```
2.36 minutes

umap5 dimensions
```{r}
start_time <- Sys.time()
set.seed(42)
umap_5D <- map(datasets_mice_app_input, function(x){
  
  custom.umap <- umap.defaults
  custom.umap$verbose <- TRUE
  custom.umap$input <- "data"
  custom.umap$n_components <- 5
  
  embedding_umap <- umap(as.matrix(tf_spread(as.data.frame(x$tf_fun))), config = custom.umap)#
  colnames(embedding_umap$layout) <- paste0("embs_umap5D_", 1:ncol(embedding_umap$layout))
  
  embedding_umap
})

#  datasets_mice_app <- map2(datasets_mice_app, tsne_noBH, ~ cbind(.x, .y$Y))
datasets_mice_app_input1 <- map2(datasets_mice_app_input1, umap_5D, ~ cbind(.x, .y$layout))

end_time <- Sys.time()
end_time - start_time
```
2.48 minutes

------------------------------------------------------------------------------
MDS

distance matrix

```{r}
# get the dist matrix (L2) for all datasets

dist_curve <- function(curve) {
  dist(tf_spread(as.data.frame(curve$tf_fun)))
}

dists <- map(datasets_mice_app_input, dist_curve)
```

plot function
```{r}
plot_gofs <- function(gofs, meth = "GOF", name = "Plot") {
 # browser()
  ggplot(as.data.frame(gofs)) + geom_point(aes(x = 1:length(gofs), y = gofs)) +
    theme_minimal() +
    xlab("number of dimensions") + ylab(meth) +
    ggtitle(name) +
    theme(plot.title = element_text(size = 9)) +
    if (meth == "GOF") geom_hline(yintercept = c(0.5, 0.75), size = 0.5, alpha = 0.5) #+
   #if (meth == "GOF") geom_hline(yintercept = 0.8, linetype = "dashed", color = "red")
    
}
```


cmdscale
```{r}
set.seed(42)
# get the MDS GOF for 1 - n-1 dims
getgof <- function(curve_dist) {
  
  message("new dataset")
  ndim <- attr(curve_dist, "Size")
  k = 9 #ndim - 1

  gofs <- numeric()
  for (dim in seq_len(k)) { 
    message(dim)
  gofs[length(gofs) + 1] <- cmdscale(as.matrix(curve_dist), k = dim, list. = TRUE)$GOF[[1]]
  }
  gofs
}

gofs_list <- map(dists[-7], getgof)

# plot the GOF per number of dimensions
map2(gofs_list, names(gofs_list), ~plot_gofs(gofs = .x, meth = "GOF", name = .y))

# at what number of dimension does the GOF surpass 0.8?
unlist(map(gofs_list, function(x) min(which(x > 0.8)))) # 5,8,6,6,6,7, - , 5

```

```{r}
gofs_all <- map(dists[7], getgof)
```

-------------------------------

MDS 5 dimensions
```{r}

start_time <- Sys.time()
set.seed(42)
MDS_5D <- map(dists[-7], function(x){
  
  message("new dataset")
  embedding_MDS <- cmdscale(x, k = 5, list. = TRUE)
  colnames(embedding_MDS$points) <- paste0("embs_MDS5D_", 1:ncol(embedding_MDS$points))
  
  embedding_MDS
})

# 
 datasets_mice_app_input1 <- map2(datasets_mice_app_input1, MDS_5D, ~ cbind(.x, .y$points))
# 

end_time <- Sys.time()
end_time - start_time
```
1.83 minutes (w/o number 7)

calculate LOF-scores on the MDS5D

```{r}

start_time <- Sys.time()
set.seed(42)
MDS_5D_lof <- map(datasets_mice_app_input2_sm, function(x){
  
  message("new dataset")
  lof_scores <- lof(x[ ,c("embs_MDS5D_1", "embs_MDS5D_2", "embs_MDS5D_3", "embs_MDS5D_4", "embs_MDS5D_5")])

  lof_scores
})

datasets_mice_app_input2_sm <- map2(datasets_mice_app_input2_sm, MDS_5D_lof, ~ cbind(.x, "vars_MDS_LOF" = .y))

end_time <- Sys.time()
end_time - start_time


```
0.65 seconds
0.33

calculate LOF-scores on the UMAP 5D

```{r}

start_time <- Sys.time()
set.seed(42)
umap_5D_lof <- map(datasets_mice_app_input2_sm, function(x){
  
  #browser()
  message("new dataset")
  lof_scores <- lof(x[ ,c("embs_umap5D_1", "embs_umap5D_2", "embs_umap5D_3", "embs_umap5D_4", "embs_umap5D_5")])

  lof_scores
})

datasets_mice_app_input2_sm <- map2(datasets_mice_app_input2_sm, umap_5D_lof, ~ cbind(.x, "vars_UMAP_LOF" = .y))

end_time <- Sys.time()
end_time - start_time


```


dataset without mice_all

```{r}
datasets_mice_app_input2 <- datasets_mice_app_input[-7]
datasets_mice_app_input2_sm <- datasets_mice_app_input1[-7]
```

sammon
```{r}
# get the dist matrix (L2) for all datasets

dists7 <- dists
dists7$mice_all <- NULL
```

```{r}

getstress_sam <- function(curve_dist) {

  ndim <- attr(curve_dist, "Size")
  k = 15

  stresses <- numeric()
  for (dim in seq_len(k)) { #ndim - 1
    message(dim)
  stresses[length(stresses) + 1] <- MASS::sammon(curve_dist, k = dim, trace = FALSE)$stress
  }
  stresses
}

stress_list <- map(dists7, getstress_sam)

save(stress_list, file = paste0(here::here(), "/stress_sammon.rds"))

# scree plots
#map(stress_list, plot_gofs, meth = "stress")
map2(stress_list, names(stress_list), ~plot_gofs(gofs = .x, meth = "stress", name = .y))


```
```{r}

start_time <- Sys.time()
set.seed(42)
MDS_5D_sammon <- map(dists7, function(x){
  
  message("new dataset")
  embedding_MDS <- MASS::sammon(x, k = 5, trace = FALSE)
  colnames(embedding_MDS$points) <- paste0("embs_MDS5D_sammon_", 1:ncol(embedding_MDS$points))
  
  embedding_MDS
})


datasets_mice_app_input2_sm <- map2(datasets_mice_app_input2_sm, MDS_5D_sammon, ~ cbind(.x, .y$points))


end_time <- Sys.time()
end_time - start_time


```
1.9 minutes

classic L1

distance matrix

```{r}
# get the dist matrix (L2) for all datasets

dist_curve_L1 <- function(curve) {
  dist(tf_spread(as.data.frame(curve$tf_fun)), method = "minkowski", p = 1)
}

dists_L1 <- map(datasets_mice_app_input2, dist_curve_L1)
```


```{r}

start_time <- Sys.time()
set.seed(42)
MDS_5D_L1 <- map(dists_L1, function(x){
  
  message("new dataset")
  embedding_MDS <- cmdscale(x, k = 5, list. = TRUE)
  colnames(embedding_MDS$points) <- paste0("embs_MDS5D_L1_", 1:ncol(embedding_MDS$points))
  
  embedding_MDS
})


datasets_mice_app_input2_sm <- map2(datasets_mice_app_input2_sm, MDS_5D_L1, ~ cbind(.x, .y$points))


end_time <- Sys.time()
end_time - start_time


```
2.01 minutes
1.96


classic canberra

distance matrix

```{r}
# get the dist matrix (L2) for all datasets

dist_curve_canb <- function(curve) {
  message("new dist")
  dist(tf_spread(as.data.frame(curve$tf_fun)), method = "canberra")
}

dists_canb <- map(datasets_mice_app_input2, dist_curve_L1)
```

```{r}

start_time <- Sys.time()
set.seed(42)
MDS_5D_canb <- map(dists_canb, function(x){
  
  message("new dataset")
  embedding_MDS <- cmdscale(x, k = 5, list. = TRUE)
  colnames(embedding_MDS$points) <- paste0("embs_MDS5D_canb_", 1:ncol(embedding_MDS$points))
  
  embedding_MDS
})


datasets_mice_app_input2_sm <- map2(datasets_mice_app_input2_sm, MDS_5D_canb, ~ cbind(.x, .y$points))


end_time <- Sys.time()
end_time - start_time

```
1.99 minutes
1.93

